Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries

### 2025-06-02 manuscript_tf/manuscript_blind.Rmd
update

---
title: |
  Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts
author:
  - name: Damien Dupré
    affil: 1
    email: damien.dupre@dcu.ie (corresponding author)
  - name: Edgar Morgenroth
    affil: 1
    email: edgar.morgenroth@dcu.ie
affiliation:
  - num: 1
    address: |
      Dublin City University Business School, Glasnevin, Dublin 9, Ireland
thanks:
  author

@@

bibliography: bibliography.bib
abstract: |
  This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups. The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM). The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa. These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
keywords: |
  COVID-19; School Closures; Intergenerational Transmission; Non-linear effects; Infection dynamics.
# \begin{jelcode}
# I18; C22.
# \end{jelcode}
output:
  bookdown::pdf_book:
    base_format: rticles::tf_article
    keep_tex: true
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
  - \usepackage{hyperref}
  - \usepackage[utf8]{inputenc}
  - \def\tightlist{}
---

@@

df_global <- read_csv(here("data/df_global.csv"))
df_global_age <- read_csv(here("data/df_global_age.csv"))
```

@@

To evaluate the impact of school closure on COVID-19 cases across different age groups, we first performed a non-linear time series regression to analyze the relationship between school closures and the number of COVID-19 cases in various age groups. In a second analysis, the transfer entropy method is employed to assess the influence of changes in case numbers between age groups, aiming to determine if school closures have secondary effects within households.

@@

@@

\begin{align*}
  \tag{1}
  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
  &\quad + f_3(closure_t, age\,group_t) + \epsilon_t
\end{align*}

\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as $\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)$ where $\phi$ is the autoregressive parameter. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.

@@

@@

The interpretation of our findings requires careful consideration of possible confounding factors. Changes in testing strategies, adherence to public health measures, and community transmission dynamics during school closures may have influenced the observed trends in COVID-19 cases. Our sensitivity analyses indicated that the overall trends remained robust after accounting for variations in testing rates and the presence of other public health interventions. However, these confounders may partially explain the heterogeneous effects observed across different age groups and countries. Future studies should explore these factors in greater detail to refine our understanding of the true impact of school closures.

In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures. While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns [@alfano2020efficacy; @molefi2021impact], prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.

@@

The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/) and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip). All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.

>>>>

---
title: |
  Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries
author:
  - name: Damien Dupré
    affil: 1
    email: damien.dupre@dcu.ie (corresponding author)
  - name: Edgar Morgenroth
    affil: 1
    email: edgar.morgenroth@dcu.ie
affiliation:
  - num: 1
    address: |
      Dublin City University Business School, Glasnevin, Dublin 9, Ireland
# thanks:
#   author

@@

bibliography: bibliography.bib
abstract: |
  The effectiveness of school closures as a COVID-19 control measure remains a subject of debate. This paper critically reassesses their impact by examining infection trends across various age groups in 12 European countries. We apply a dual analytical approach: Generalised Additive Models (GAMs) are used to capture complex, non-linear effects of closures on age-specific case numbers, while Transfer Entropy (TE) is employed to quantify directional transmission patterns between these age groups. Our findings reveal deviations from commonly held assumptions. While school closures were associated with a non-linear decline in overall national COVID-19 cases, the effects on children and young adults varied considerably. A consistent downward trend in infections was seen only in the pre-school age group. In contrast, school-aged children experienced a marked rise in COVID-19 cases following an initial period of stability or slight decrease after closures began. The Transfer Entropy analysis also identified asymmetric, directional patterns in transmission, revealing that infection trends in specific age groups could predict subsequent changes in others. These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.

keywords: |
  COVID-19; School Closures; Intergenerational Transmission; Non-linear effects; Infection dynamics.
# \begin{jelcode}
# I18; C22.
# \end{jelcode}
output:
  bookdown::pdf_book:
    base_format: rticles::tf_article
    keep_tex: true
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \floatplacement{figure}{H}
  - \floatplacement{table}{H}
  - \usepackage{caption}
  - \usepackage{hyperref}
  - \usepackage[utf8]{inputenc}
  - \def\tightlist{}
---

@@

# df_global <- read_csv(here("data/df_global.csv"))
# df_global_age <- read_csv(here("data/df_global_age.csv"))

```

@@

A gap in these previous analyses is the insufficient granularity in examining age-specific effects and the complex dynamics of inter-age group transmission. This study aims to address these deficiencies by providing a more robust and nuanced assessment. First, Generalized Additive Models (GAMs) are utilized to flexibly model and quantify the potentially non-linear evolution of COVID-19 cases following school closures, specifically within these fine-grained age brackets. This allows us to move beyond simple pre-post comparisons and capture dynamic temporal patterns. Then, Transfer Entropy (TE), an information-theoretic measure, is applied to assess directional causality in time series data. TE is particularly adept at identifying how past states of one variable (e.g., cases in one age group) influence future states of another (e.g., cases in a different age group), even in complex, non-linear systems like epidemic spread. This dual methodology allows for a deeper understanding of both the direct age-specific consequences of school closures and the potential for altered intergenerational transmission dynamics. Our findings challenge simplistic narratives about school closure efficacy and highlight the importance of age-stratified analysis for future pandemic preparedness.

@@

## Controlling for confounding Non-Pharmaceutical Interventions

Recognizing that school closures rarely occurred in isolation, a critical component of this analysis was to account for other contemporaneous NPIs. To mitigate confounding from other NPIs, we incorporated four time-varying composite NPI index identified by the Oxford COVID-19 Government Response Tracker (OxCGRT) project [@hale2021global]: the average stringency index, the average government response index, the average containment health index, and the average economic support index. Specific data were collected from each country on various policy responses governments implemented to curb the spread of COVID-19, such as travel restrictions, workplace closures, public event cancellations, restrictions on gatherings, or public health messaging. These indicators were then aggregated to create indexes, reflecting the overall level of government restrictions.

@@

\begin{align*}
  \tag{1}
  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
  &\quad + f_3(closure_t, age\,group_t) + f_4(stringency\,index_t) \\
  &\quad + f_5(government\,response\,index_t) + f_6(containment\,health\,index_t) \\
  &\quad + f_7(economic\,support\,index_t) +\epsilon_t
\end{align*}

\noindent where $n\prime_t$ represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting [@loader2006local], and $t$ is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms $f_1$ to $f_3$ are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The terms $f_4$ to $f_7$ are random effects corresponding to the average stringency index, the average government response index, the average containment health index, and the average economic support index. The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as $\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)$ where $\phi$ is the autoregressive parameter. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data’s temporal dynamics.

@@

While the general patterns described above were evident in the aggregated analysis, country-specific smooths indicated some national variations in the timing and magnitude of these age-specific effects, although the paradoxical increase in the 5-14 age cohorts was observed in a majority of countries.

@@

This study, despite its methodological strengths, is subject to certain limitations that warrant consideration. Foremost among these is the inherent challenge of perfectly isolating the effect of school closures from other simultaneously implemented and dynamically changing NPIs. While we incorporated four composite NPI index derived from the Oxford COVID-19 Government Response Tracker (OxCGRT) project to account for the broader public health context, residual confounding from unmeasured aspects of NPIs or behavioural changes may still persist. Our sensitivity analyses showed consistent directional effects for school closures even under varying overall NPI indexes provide some confidence, but this remains an inherent complexity in observational studies of this nature. Another consideration is the ecological design of the study, which uses aggregated country-level data; this may mask important sub-national variations and precludes inferences about individual-level risk factors. In addition, the definition of "closure" which excluded "partially open" school statuses to simplify the intervention definition, means the impact of such intermediate states was not captured. Variations in testing strategies, case definitions, and reporting practices across countries and over time could also have influenced the observed case counts, despite efforts to use standardized data sources. Lastly, this study predominantly covers periods before widespread COVID-19 vaccination and the emergence of later variants such as Omicron; the effects of school closures might differ under those altered epidemiological conditions.

Future studies should look more closely at why the virus spread between age groups as it did, and why TE patterns differed by country. This could involve using information on people's behaviour or more detailed data on other health rules. It would also be useful to study how different virus types and vaccination levels change these effects. Learning more about other influencing factors will help us better understand the real impact of school closures.

# Conclusion

This comprehensive analysis of COVID-19 school closures across 12 European nations reveals a complex and often counter-intuitive reality that challenges prevailing assumptions about their universal efficacy, particularly for children aged 5 and older. While national-level data showed an overall decrease in cases during closures, our age-disaggregated analysis using GAMs demonstrated that school-aged children (5-14 years) and young adults (15-24 years) frequently experienced periods of increased infections post-closure, a stark contrast to the consistent reductions seen only in pre-schoolers. Furthermore, Transfer Entropy analysis highlighted that the directional influence of transmission between age groups was asymmetric and varied by country, suggesting that the role of children in broader community transmission during such periods is not uniform.

These findings underscore the critical importance of nuanced, age-specific analysis in evaluating public health interventions and caution against the broad application of measures like school closures without robust evidence of their benefit across all targeted demographics and a clear understanding of potential unintended consequences [@alfano2020efficacy; @molefi2021impact]. For future pandemic preparedness, a more targeted, evidence-based approach is essential, prioritising measures that minimise societal disruption while effectively protecting public health across all age strata.

@@

The data can be accessed from the COVerAGE-DB OSF repository (https://osf.io/mpwjq/), from the Oxford Covid-19 Government Response Tracker (OxCGRT) "covid-policy-dataset" GitHub repository (https://github.com/OxCGRT/covid-policy-dataset), and from the UNESCO servers (https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip). All preprocessing, analyses and code used to build the submitted paper are available at https://github.com/damien-dupre/covid_ts_causality.

### 2025-06-02 manuscript_tf/manuscript_blind.tex
update

\title{Examining the Impact of School Closures on COVID-19 Infections in Europe and their Effects on Different Age Cohorts}

@@

\thanks{CONTACT Damien Dupré. Email: \href{mailto:damien.dupre@dcu.ie}{\nolinkurl{damien.dupre@dcu.ie}} (corresponding author), Edgar Morgenroth. Email: \href{mailto:edgar.morgenroth@dcu.ie}{\nolinkurl{edgar.morgenroth@dcu.ie}}}

@@

\begin{abstract}
This paper analyzes the trends of COVID-19 cases in different age groups in selected European countries, assessing to what extent school closures helped reduce case numbers, how school closures affected case numbers in different age groups, and how COVID-19 spread between age groups. The study also evaluates the effects of school closures on COVID-19 cases using a Generalized Additive Model (GAM). The findings demonstrate a decreasing non-linear effect of school closure on the total COVID-19 case number across all countries studied. However, the analysis of age groups only reveals a consistent downward trend in infections in the 0 to 4 pre-school age group, while the school-going age groups 5 to 14 exhibit a significant increase in cases. Age groups 15 to 24 show a surge immediately after closure, followed by a decline. Transfer Entropy calculations highlight asymmetry in age group influences, indicating that changes in COVID-19 cases in certain age groups predict changes in other age groups but not vice versa. These findings contribute to a better understanding of COVID-19 dynamics in European countries and provide insights for public health strategies and interventions.
\end{abstract}

@@

\begin{jelcode}
I18; C22.
\end{jelcode}

@@

@@

\begin{align*}
  \tag{1}
  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
  &\quad + f_3(closure_t, age\,group_t) + \epsilon_t
\end{align*}

\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting \citep{loader2006local}, and \(t\) is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms \(f_1\) to \(f_3\) are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as \(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\) where \(\phi\) is the autoregressive parameter. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.

@@

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB \citep{riffe2021data}.}\label{fig:overall}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions. Source: \citet{unesco2022https}.}\label{fig:descriptive}
\end{figure}

As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old. The obtained results satisfy the requirements to fit this model, which explains 77.3\% of the variation in COVID-19 cases.

Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.55) = 3302.08\), \(p < 0.001\); Belgium: \(\chi^2(5.99) = 7680.99\), \(p < 0.001\); Bulgaria: \(\chi^2(5.55) = 125.19\), \(p < 0.001\); Croatia: \(\chi^2(5.6) = 734.51\), \(p < 0.001\); Estonia: \(\chi^2(5.88) = 2157.39\), \(p < 0.001\); France: \(\chi^2(4.98) = 4141.45\), \(p < 0.001\); Germany (\(\chi^2(5.97) = 3058.7\), \(p < 0.001\)); Greece: \(\chi^2(5.9) = 1793.54\), \(p < 0.001\); The Netherlands: \(\chi^2(5.99) = 7108.09\), \(p < 0.001\); Portugal: \(\chi^2(5.81) = 3104.43\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 6218.66\), \(p < 0.001\); Spain: \(\chi^2(5.95) = 559.14\), \(p < 0.001\), see Figure \ref{fig:age}).

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.}\label{fig:country}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups.}\label{fig:age}
\end{figure}

\begin{figure}[H]
\includegraphics[width=\textwidth]{manuscript_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country. Age groups on the x-axis are influencing the age groups on the y-axis (\(ag\,x \rightarrow\,ag\,y\)). The significance of each Transfer Entropy coefficient is provided in Appendix 2.}\label{fig:te}
\end{figure}

Considering all countries, the analysis of age groups reveals distinct patterns (Figure \ref{fig:te}). For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (\(\chi^2(5.94) = 590.81\), \(p < 0.001\)). A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (\(\chi^2(5.96) = 1186.05\), \(p < 0.001\)), where a decrease is observed during the first two weeks, followed by an increase. while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (\(\chi^2(5.97) = 307.56\), \(p < 0.001\)). For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(1.01) = 57.27\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(4.95) = 5624.5\), \(p < 0.001\)). Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.

@@

The interpretation of our findings requires careful consideration of possible confounding factors. Changes in testing strategies, adherence to public health measures, and community transmission dynamics during school closures may have influenced the observed trends in COVID-19 cases. Our sensitivity analyses indicated that the overall trends remained robust after accounting for variations in testing rates and the presence of other public health interventions. However, these confounders may partially explain the heterogeneous effects observed across different age groups and countries. Future studies should explore these factors in greater detail to refine our understanding of the true impact of school closures.

In conclusion, this study sheds light on the trends and patterns of confirmed COVID-19 cases in selected European countries, with a particular focus on the influence of age groups and school closures. While a reduction in overall COVID-19 cases was found during school closures, which might suggest that like lockdowns \citep{alfano2020efficacy, molefi2021impact}, prolonged school closures could contribute to mitigating the spread of the virus, the results for the school-going age groups suggest that this is not the case for all age groups.

@@

The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}) and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}). All preprocessing, analyses and code used to build the submitted paper are available at \url{https://github.com/damien-dupre/covid_ts_causality}.

>>>>

\title{Divergent Impacts of COVID-19 School Closures on Youth Infection Dynamics: Evidence from 12 European Countries}

@@

@@

\begin{abstract}
The effectiveness of school closures as a COVID-19 control measure remains a subject of debate. This paper critically reassesses their impact by examining infection trends across various age groups in 12 European countries. We apply a dual analytical approach: Generalised Additive Models (GAMs) are used to capture complex, non-linear effects of closures on age-specific case numbers, while Transfer Entropy (TE) is employed to quantify directional transmission patterns between these age groups. Our findings reveal deviations from commonly held assumptions. While school closures were associated with a non-linear decline in overall national COVID-19 cases, the effects on children and young adults varied considerably. A consistent downward trend in infections was seen only in the pre-school age group. In contrast, school-aged children experienced a marked rise in COVID-19 cases following an initial period of stability or slight decrease after closures began. The Transfer Entropy analysis also identified asymmetric, directional patterns in transmission, revealing that infection trends in specific age groups could predict subsequent changes in others. These results challenge the assumption that school closures uniformly benefit younger populations and emphasise the need for age-specific analysis in pandemic planning and support the development of more nuanced, evidence-based public health policies.
\end{abstract}

@@

@@

\subsection{Controlling for confounding Non-Pharmaceutical Interventions}\label{controlling-for-confounding-non-pharmaceutical-interventions}

Recognizing that school closures rarely occurred in isolation, a critical component of this analysis was to account for other contemporaneous NPIs. To mitigate confounding from other NPIs, we incorporated four time-varying composite NPI index identified by the Oxford COVID-19 Government Response Tracker (OxCGRT) project \citep{hale2021global}: the average stringency index, the average government response index, the average containment health index, and the average economic support index. Specific data were collected from each country on various policy responses governments implemented to curb the spread of COVID-19, such as travel restrictions, workplace closures, public event cancellations, restrictions on gatherings, or public health messaging. These indicators were then aggregated to create indexes, reflecting the overall level of government restrictions.

@@

\begin{align*}
  \tag{1}
  n\prime_t &\sim \text{Poisson}(\lambda_t) \\
  \log(\lambda_t) &= \beta_0 + \beta_1\,wave_t + \beta_2\,country_t + \beta_3\,age\,group_t \\
  &\quad + f_1(closure_t) + f_2(closure_t, country_t) \\
  &\quad + f_3(closure_t, age\,group_t) + f_4(stringency\,index_t) \\
  &\quad + f_5(government\,response\,index_t) + f_6(containment\,health\,index_t) \\
  &\quad + f_7(economic\,support\,index_t) +\epsilon_t
\end{align*}

\noindent where \(n\prime_t\) represents the confirmed COVID-19 cases, assuming a Poisson distribution for the fitting \citep{loader2006local}, and \(t\) is the date corresponding to the confirmed COVID-19 cases. The response variable includes a specific random effect taking into account variation within waves of school closure, countries, and age groups. The terms \(f_1\) to \(f_3\) are smooth functions of the time since closure, the time since closure for each country, and the time since closure for each age group. The terms \(f_4\) to \(f_7\) are random effects corresponding to the average stringency index, the average government response index, the average containment health index, and the average economic support index. The restricted maximum likelihood (REML) was used to avoid over fitting while estimating smoothing parameters. In order to accurately account for the autocorrelation arising from the time series data, the residuals are modelled using an AR1 error model such as \(\epsilon_t = \phi \epsilon_{t-1} + \eta_t, \quad \eta_t \sim \mathcal{N}(0, \sigma^2)\) where \(\phi\) is the autoregressive parameter. By incorporating the autoregressive component, the AR1 model acknowledges the dependence of each residual on its previous value, thus providing a comprehensive representation of the data's temporal dynamics.

@@

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/overall-1} \caption{Cumulative COVID-19 cases number for selected European countries since the beginning of the pandemic. Source: COVerAGE-DB \citep{riffe2021data}.}\label{fig:overall}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/descriptive-1} \caption{Periods of school closure since the beginning of the COVID-19 pandemic for selected European countries and their reason: regular academic break versus closure due to government decisions. Source: \citet{unesco2022https}.}\label{fig:descriptive}
\end{figure}

As outlined above, to evaluate the shape of the trend in the numbers of COVID-19 cases reported after the three school closures longer than 21 consecutive days, a GAM was fitted as described above, taking into account the overall effect across all the selected European countries, as well as the effect for age groups: 0 to 4, 5 to 9, 10 to 14, 15 to 19, and 20 to 24 years old. The obtained results satisfy the requirements to fit this model, which explains 80.2\% of the variation in COVID-19 cases.

Overall, the results revealed a decreasing non-linear effect of school closure at a country level for the selected European countries (Austria: \(\chi^2(5.82) = 3243.01\), \(p < 0.001\); Belgium: \(\chi^2(5.98) = 6206.45\), \(p < 0.001\); Bulgaria: \(\chi^2(5.89) = 258.19\), \(p < 0.001\); Croatia: \(\chi^2(5.1) = 740.24\), \(p < 0.001\); Estonia: \(\chi^2(5.93) = 1834.26\), \(p < 0.001\); France: \(\chi^2(1) = 2412.59\), \(p < 0.001\); Germany (\(\chi^2(5.98) = 3081.05\), \(p < 0.001\)); Greece: \(\chi^2(5.95) = 2423.67\), \(p < 0.001\); The Netherlands: \(\chi^2(4.97) = 5878.92\), \(p < 0.001\); Portugal: \(\chi^2(5.94) = 2692.56\), \(p < 0.001\); Slovakia (\(\chi^2(5.97) = 5335.74\), \(p < 0.001\); Spain: \(\chi^2(5.94) = 498.09\), \(p < 0.001\), see Figure \ref{fig:age}).

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/country-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by country. Standardized effects are reported to compare the shape of the curve between countries.}\label{fig:country}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/age-1} \caption{Standardized effect of the smooth term in Generalized Additive Model by age group. Standardized effects are reported to compare the shape of the curve between age groups.}\label{fig:age}
\end{figure}

While the general patterns described above were evident in the aggregated analysis, country-specific smooths indicated some national variations in the timing and magnitude of these age-specific effects, although the paradoxical increase in the 5-14 age cohorts was observed in a majority of countries.

\begin{figure}[H]
\includegraphics[width=\textwidth]{manuscript_blind_files/figure-latex/te-1} \caption{Matrix of Transfer Entropy coefficients according to every age group combination for each of the selected European country. Age groups on the x-axis are influencing the age groups on the y-axis (\(ag\,x \rightarrow\,ag\,y\)). The significance of each Transfer Entropy coefficient is provided in Appendix 2.}\label{fig:te}
\end{figure}

Considering all countries, the analysis of age groups reveals distinct patterns (Figure \ref{fig:te}). For the 0 to 4 age group, i.e., the pre-school group, there was a downward trend during the initial three weeks following school closure, followed by an increase during the fourth week (\(\chi^2(5.96) = 410.48\), \(p < 0.001\)). A similar, but more moderate, trend is observer for the age group ranging from 5 to 9 (\(\chi^2(5.95) = 878.29\), \(p < 0.001\)), where a decrease is observed during the first two weeks, followed by an increase. while the trend for the age group ranging from 10 to 14 is flat during the first two weeks, a similar increase is shown at the beginning of the third week of school closure (\(\chi^2(1) = 61.62\), \(p < 0.001\)). For age groups between 15 and 24, there is a notable surge immediately after school closure, followed by a decrease after 14 days (15 to 19 age group: \(\chi^2(5.97) = 7929.88\), \(p < 0.001\) and 20 to 24 age group: \(\chi^2(4.98) = 18386.65\), \(p < 0.001\)). Thus, while the total case number declined following school closures, among younger age groups only the pre-school age group had a consistent reduction in cases, while for other age groups, there was even an increase in cases following school closures.

@@

This study, despite its methodological strengths, is subject to certain limitations that warrant consideration. Foremost among these is the inherent challenge of perfectly isolating the effect of school closures from other simultaneously implemented and dynamically changing NPIs. While we incorporated four composite NPI index derived from the Oxford COVID-19 Government Response Tracker (OxCGRT) project to account for the broader public health context, residual confounding from unmeasured aspects of NPIs or behavioural changes may still persist. Our sensitivity analyses showed consistent directional effects for school closures even under varying overall NPI indexes provide some confidence, but this remains an inherent complexity in observational studies of this nature. Another consideration is the ecological design of the study, which uses aggregated country-level data; this may mask important sub-national variations and precludes inferences about individual-level risk factors. In addition, the definition of ``closure'' which excluded ``partially open'' school statuses to simplify the intervention definition, means the impact of such intermediate states was not captured. Variations in testing strategies, case definitions, and reporting practices across countries and over time could also have influenced the observed case counts, despite efforts to use standardized data sources. Lastly, this study predominantly covers periods before widespread COVID-19 vaccination and the emergence of later variants such as Omicron; the effects of school closures might differ under those altered epidemiological conditions.

Future studies should look more closely at why the virus spread between age groups as it did, and why TE patterns differed by country. This could involve using information on people's behaviour or more detailed data on other health rules. It would also be useful to study how different virus types and vaccination levels change these effects. Learning more about other influencing factors will help us better understand the real impact of school closures.

\section{Conclusion}\label{conclusion}

This comprehensive analysis of COVID-19 school closures across 12 European nations reveals a complex and often counter-intuitive reality that challenges prevailing assumptions about their universal efficacy, particularly for children aged 5 and older. While national-level data showed an overall decrease in cases during closures, our age-disaggregated analysis using GAMs demonstrated that school-aged children (5-14 years) and young adults (15-24 years) frequently experienced periods of increased infections post-closure, a stark contrast to the consistent reductions seen only in pre-schoolers. Furthermore, Transfer Entropy analysis highlighted that the directional influence of transmission between age groups was asymmetric and varied by country, suggesting that the role of children in broader community transmission during such periods is not uniform.

These findings underscore the critical importance of nuanced, age-specific analysis in evaluating public health interventions and caution against the broad application of measures like school closures without robust evidence of their benefit across all targeted demographics and a clear understanding of potential unintended consequences \citep{alfano2020efficacy, molefi2021impact}. For future pandemic preparedness, a more targeted, evidence-based approach is essential, prioritising measures that minimise societal disruption while effectively protecting public health across all age strata.

@@

The data can be accessed from the COVerAGE-DB OSF repository (\url{https://osf.io/mpwjq/}), from the Oxford Covid-19 Government Response Tracker (OxCGRT) ``covid-policy-dataset'' GitHub repository (\url{https://github.com/OxCGRT/covid-policy-dataset}), and from the UNESCO servers (\url{https://en.unesco.org/file/unesco-data-school-closures-february-2020-june-2022csv-zip}). All preprocessing, analyses and code used to build the submitted paper are available at \url{https://github.com/damien-dupre/covid_ts_causality}.
